---
title: '`R` Shiny'
subtitle: 'building web application with Shiny in `R`'
author: 'Bonnie Cooper'
output:
  rmdformats::downcute
---

```{r message=F}
library( DiagrammeR )
library( shiny )
library( plotly )
library( babynames )
library( shinythemes )
library( gapminder )
library( dplyr )
```


Introduction to Shiny apps in `R`

What is a web app?  

* Updates based on user input/interaction
* Made up of UI & server  

How does a web app work?  
A web app is a thingy that updates based on user input/interaction. Most web application consist of two parts. The client contains the user interface, that is, buttons and selectors and text boxes and other things that the user can interact with. The server (or backend) is where computation happens, including things like manipulating data and running models.  

1. Client (User Interface)
2. Server (Backend) that carries out computations based on the user interactions
```{r, message=F, echo=F}
DiagrammeR::grViz("digraph {

graph [layout = dot, rankdir = LR]

# define the global styles of the nodes. We can override these in box if we wish
node [shape = rectangle, style = filled, fillcolor = Linen]

data1 [label = 'Client 1', shape = folder, fillcolor = Beige]
data2 [label = 'Client 2', shape = folder, fillcolor = Beige]
data3 [label = 'Client 2', shape = folder, fillcolor = Beige]
process [label =  'teh Internetz']
statistical [label = 'Server']

# edge definitions with the node IDs
{data1 data2 data3}  -> process -> statistical
}")
```


`iris` shiny app
```{r}

#plot_kmean()

ui <- fluidPage(
  h1( 'K-Means Clustering App' ),
  selectInput( 'x', 'Select x', names( iris ), 'Sepal.Length' ),
  selectInput( 'y', 'Select y', names( iris ), 'Sepal.Width' ),
  numericInput('nb_clusters', 'Select number of clusters', 3 ),
  plotly::plotlyOutput( 'kmeans_plot' )
)

server <- function( input, output, session ){
  plot_kmeans <- function( data, x, y, nb_clusters ){
  k1 <- kmeans(x=data[, 1:4], centers=nb_clusters)
  plot( data[ c(x,y) ], col = k1$cluster )}
  output$kmeans_plot <- plotly::renderPlotly({
    plot_kmeans(iris, input$x, input$y, input$nb_clusters)
  })
}

shinyApp( ui = ui, server = server )
```


Shiny skeleton  

* Load `shiny`
* Create the UI with an HTML function
  + `fluidPage()`
* Define a custom function to create the server
  + ex: input, output & session
* Run the app
  + `shinyApp()`
  
  Hello Shiny!
```{r message = F}
ui <- fluidPage(
  "Hello Shiny!"
)
server <- function( input, output, session ){
  
}
shinyApp( ui = ui, server = server )
```
  
Ask a Question:
```{r message = F}
ui <- fluidPage(
  textInput( 'name', 'Enter a name:'),
  textOutput("q")
)
server <- function( input, output ){
  output$q <- renderText( {
    paste( "Do you prefer dogs or cats,",
           input$name, "?")
  })
}
shinyApp( ui = ui, server = server )
```

```{r}
data( babynames )
ui <- fluidPage(
  titlePanel( 'Baby Name Explorer' ),
  sidebarLayout( sidebarPanel( 
    textInput( 'name', 'Enter Name', 'David' ) ),
  mainPanel( plotOutput( 'trend' ) ) )
)
server <- function( input, output, session ){
  output$trend <- renderPlot( {
    data_name <- subset(
      babynames, name == input$name
    )
    ggplot( data_name ) +
      geom_line(
        aes( x = year, y = prop, color = sex )
      )
  })
}
shinyApp( ui = ui, server = server )
```

### Example Inputs

* slider - slide a pointer to a value on a scale
* select list - a pull down menu allows the user to select between options on a list
* numeric - can enter or increment integer values
* date range - allows user to select a range of dates

```{r}
ui <- fluidPage(
  titlePanel("What's in a Name?"),
  # Add select input named "sex" to choose between "M" and "F"
  selectInput('sex', 'Select Sex', choices = c("F", "M")),
  # CODE BELOW: Add slider input named 'year' to select years  (1900 - 2010)
  sliderInput( "year", "label", value = 1900, min = 1900, max = 2010 ),
  # Add plot output to display top 10 most popular names
  plotOutput('plot_top_10_names')
)

server <- function(input, output, session){
  # Render plot of top 10 most popular names
  output$plot_top_10_names <- renderPlot({
    # Get top 10 names by sex and year
    top_10_names <- babynames %>% 
      filter(sex == input$sex) %>% 
    # MODIFY CODE BELOW: Filter for the selected year
      filter(year == input$year ) %>% 
      top_n(10, prop)
    # Plot top 10 names by sex and year
      ggplot(top_10_names, aes(x = name, y = prop)) +
        geom_col(fill = "#263e63")
  })
}
shinyApp( ui = ui, server = server )
```

```{r}
ui <- fluidPage(
  titlePanel("What's in a Name?"),
  # CODE BELOW: Add select input named "sex" to choose between "M" and "F"
  selectInput( "sex", "Male or Female?", choices = c( "M", "F" ) ),
  # Add plot output to display top 10 most popular names
  plotOutput('plot_top_10_names')
)

server <- function(input, output, session){
  # Render plot of top 10 most popular names
  output$plot_top_10_names <- renderPlot({
    # Get top 10 names by sex and year
    top_10_names <- babynames %>% 
      # MODIFY CODE BELOW: Filter for the selected sex
      filter(sex == input$sex ) %>% 
      filter(year == 1900) %>% 
      top_n(10, prop)
    # Plot top 10 names by sex and year
    ggplot(top_10_names, aes(x = name, y = prop)) +
      geom_col(fill = "#263e63")
  })
}

shinyApp(ui = ui, server = server)
```


### Render & Output Functions

Render functions build outputs in the server based on inputs  
Types of render functions:  

* renderText()
* renderTable()
* renderImage()
* renderPlot()
* check out the shiny documentation for more

Output functions are in the ui to visualize the result of the render functions in the server
Types of output function:  

* textOutput()
* tableOutput()
* imageOutput()
* plotOutput()
* dataTableOutput()

```{r}
ui <- fluidPage(
  DT::DTOutput( "babynames_table" )
)

server <- function( input, output ){
  output$babynames_table <- DT::renderDT({
    babynames %>%
      dplyr::sample_frac(.1)
  })
}

shinyApp(ui = ui, server = server)
```

```{r}
ui <- fluidPage(
  titlePanel("What's in a Name?"),
  # Add select input named "sex" to choose between "M" and "F"
  selectInput('sex', 'Select Sex', choices = c("F", "M")),
  # Add slider input named "year" to select year between 1900 and 2010
  sliderInput('year', 'Select Year', min = 1900, max = 2010, value = 1900),
  # CODE BELOW: Add table output named "table_top_10_names"
  DT::DTOutput( "table_top_10_names" )
  
)
server <- function(input, output, session){
  # Function to create a data frame of top 10 names by sex and year 
  top_10_names <- function(){
    top_10_names <- babynames %>% 
      filter(sex == input$sex) %>% 
      filter(year == input$year) %>% 
      top_n(10, prop)
  }
  # CODE BELOW: Render a table output named "table_top_10_names"
  output$table_top_10_names <- DT::renderDT({
    top_10_names() })
}
shinyApp(ui = ui, server = server)
```

```{r}

name <- c( 'Kizzy', 'Deneen', 'Royalty', 'Mareli', 'Moesha', 'Marely', 'Kanye', 
           'Tennille', 'Aitana', 'Kadijah','Shaquille', 'Catina', 'Allisson', 
           'Emberly', 'Nakia' , 'Jaslene', 'Kyrie', 'Akeelah', 'Zayn', 'Talan' )
sex <- c( 'F', 'F', 'F', 'F', 'F', 'F', 'M', 'F', 'F', 'F', 'M', 'F', 'F', 
          'F', 'M', 'F', 'M', 'F', 'M', 'M' )
total <- c(2325, 3603, 1806, 1024, 1067, 2577, 1319, 2172, 1625, 1418, 5439, 
           4178, 2377, 1471, 1991, 2870, 5858, 1331, 3347, 3640 )
max <- c( 1116, 1604, 747, 411, 426, 1004, 508, 769, 564, 486, 1784, 1370, 767, 
          467, 612, 872, 1774, 403, 988, 1059 )
nb_years <- c( 30, 52, 14, 21, 14, 28, 16, 32, 23, 36, 29, 47, 21, 34, 40, 17, 31, 17, 25, 28 )
trendiness <- c( 0.48, 0.445, 0.414, 0.401, 0.399, 0.390, 0.385, 0.354, 0.347, 
                 0.343, 0.328, 0.328, 0.323, 0.317, 0.307, 0.304, 0.303, 0.303, 
                 0.295, 0.291 )
top_trendy_names <- data.frame( 'name' = name, 'sex' = sex, 
                                'total' = total, 'max' = max,
                                'nb_years' = nb_years, 'trendiness' = trendiness )  
str( top_trendy_names )
```

top_trendy_names
 A tibble: 20 x 6
   name      sex   total   max nb_years trendiness
   <chr>     <chr> <int> <int>    <int>      <dbl>
 1 Kizzy     F      2325  1116       30      0.48 
 2 Deneen    F      3603  1604       52      0.445
 3 Royalty   F      1806   747       14      0.414
 4 Mareli    F      1024   411       21      0.401
 5 Moesha    F      1067   426       14      0.399
 6 Marely    F      2577  1004       28      0.390
 7 Kanye     M      1319   508       16      0.385
 8 Tennille  F      2172   769       32      0.354
 9 Aitana    F      1625   564       23      0.347
10 Kadijah   F      1418   486       36      0.343
11 Shaquille M      5439  1784       29      0.328
12 Catina    F      4178  1370       47      0.328
13 Allisson  F      2377   767       21      0.323
14 Emberly   F      1471   467       34      0.317
15 Nakia     M      1991   612       40      0.307
16 Jaslene   F      2870   872       17      0.304
17 Kyrie     M      5858  1774       31      0.303
18 Akeelah   F      1331   403       17      0.303
19 Zayn      M      3347   988       25      0.295
20 Talan     M      3640  1059       28      0.291

```{r}
ui <- fluidPage(
  selectInput('name', 'Select Name', top_trendy_names$name),
  # CODE BELOW: Add a plotly output named 'plot_trendy_names'
  plotly::plotlyOutput( 'plot_trendy_names' )
)
server <- function(input, output, session){
  # Function to plot trends in a name
  plot_trends <- function(){
     babynames %>% 
      filter(name == input$name) %>% 
      ggplot(aes(x = year, y = n)) +
      geom_col()
  }
  # CODE BELOW: Render a plotly output named 'plot_trendy_names'
  output$plot_trendy_names <- plotly::renderPlotly( {plot_trends()} )
}
shinyApp(ui = ui, server = server)
```
 
### Layouts and Themes
well chosen layout are good for aesthetics!  

Default:
```{r}
ui <- fluidPage(
  titlePanel( "histogram" ),
  sliderInput( "nb_bins", "# Bins", 5, 10, 5 ),
  plotOutput( "hist" )
)
server <- function( input, output, session ){
  output$hist <- renderPlot( {
    hist( faithful$waiting,
          breaks = input$nb_bins,
          col = 'pink' )
  })
}
shinyApp( ui = ui, server = server )
```

Sidebar layout: inputs to the left, output in the main panel to the right
```{r}
ui <- fluidPage(
  titlePanel( "histogram" ),
  sidebarLayout( sidebarPanel( sliderInput( "nb_bins", "# Bins", 5, 10, 5 ) ),
  mainPanel( plotOutput( "hist" ) ) )
)
server <- function( input, output, session ){
  output$hist <- renderPlot( {
    hist( faithful$waiting,
          breaks = input$nb_bins,
          col = 'pink' )
  })
}
shinyApp( ui = ui, server = server )
```
Tab Layout  
put different plots in different tabs to give each it's own space
```{r}
ui <- fluidPage(
  titlePanel( 'Histogram' ),
  sidebarLayout(
    sidebarPanel( sliderInput( 'nb_bins', '# Bins', 5, 10, 5 ) ),
    mainPanel( 
      tabsetPanel(
        tabPanel( 'Waiting', plotOutput( 'hist_waiting' ) ),
        tabPanel( 'Eruptions', plotOutput( 'hist_eruptions' ) )
      ))
  )
)
server <- function( input, output, session ){
  output$hist_waiting <- renderPlot( {
    hist(faithful$waiting,
         breaks = input$nb_bins,
         col = 'pink' )
  })
}
shinyApp( ui = ui, server = server )
```
Theme selector  
Shiny makes it easy to customize the theme of an app. The UI functions in Shiny make use of Twitter Bootstrap, a popular framework for building web applications. Bootswatch extends Bootstrap by making it really easy to skin an application with minimal code changes.
```{r}
ui <- fluidPage(
  titlePanel( "histogram" ),
  #shinythemes::themeSelector(),
  theme = shinythemes::shinytheme( 'superhero' ),
  sidebarLayout( sidebarPanel( sliderInput( "nb_bins", "# Bins", 5, 10, 5 ) ),
  mainPanel( plotOutput( "hist" ) ) )
)
server <- function( input, output, session ){
  output$hist <- renderPlot( {
    hist( faithful$waiting,
          breaks = input$nb_bins,
          col = 'pink' )
  })
}
shinyApp( ui = ui, server = server )
```

```{r}
ui <- fluidPage(
  # MODIFY CODE BELOW: Wrap in a sidebarLayout
    # MODIFY CODE BELOW: Wrap in a sidebarPanel
    sidebarLayout( sidebarPanel( selectInput('name', 'Select Name', top_trendy_names$name) ),
    # MODIFY CODE BELOW: Wrap in a mainPanel
    mainPanel( plotly::plotlyOutput('plot_trendy_names') ,
    DT::DTOutput('table_trendy_names') ) )
)
# DO NOT MODIFY
server <- function(input, output, session){
  # Function to plot trends in a name
  plot_trends <- function(){
     babynames %>% 
      filter(name == input$name) %>% 
      ggplot(aes(x = year, y = n)) +
      geom_col()
  }
  output$plot_trendy_names <- plotly::renderPlotly({
    plot_trends()
  })
  
  output$table_trendy_names <- DT::renderDT({
    babynames %>% 
      filter(name == input$name)
  })
}
shinyApp(ui = ui, server = server)
```
Tab Example:
```{r}
ui <- fluidPage(
  sidebarLayout(
    sidebarPanel(
      selectInput('name', 'Select Name', top_trendy_names$name)
    ),
    mainPanel(
      # MODIFY CODE BLOCK BELOW: Wrap in a tabsetPanel
      tabsetPanel(
        # MODIFY CODE BELOW: Wrap in a tabPanel providing an appropriate label
        tabPanel( 'Plot', plotly::plotlyOutput('plot_trendy_names') ),
        # MODIFY CODE BELOW: Wrap in a tabPanel providing an appropriate label
        tabPanel( 'Table', DT::DTOutput('table_trendy_names') ) )
    )
  )
)
server <- function(input, output, session){
  # Function to plot trends in a name
  plot_trends <- function(){
     babynames %>% 
      filter(name == input$name) %>% 
      ggplot(aes(x = year, y = n)) +
      geom_col()
  }
  output$plot_trendy_names <- plotly::renderPlotly({
    plot_trends()
  })
  
  output$table_trendy_names <- DT::renderDT({
    babynames %>% 
      filter(name == input$name)
  })
}
shinyApp(ui = ui, server = server)
```


### Building apps
Build an app using the gapminder dataset.  
Explore Life Expectancy vs. GDP per Capita  

Building Shiny apps: 4 steps  

1. Add inputs (UI)
2. Add outputs (server)
3. Update layout (UI)
4. Update outputs (Server)

```{r}
ui <- fluidPage(
  titlePanel( 'Life Expectation vs. GDP Per Capita' ), #step1
  selectInput( 'continent', 'Select Continent', unique( gapminder$continent )), #step1
  sliderInput( 'year', 'Select Year', 1952, 2007, 1992, step = 5 ), #step1
  plotOutput( 'plot' ), #step2
  DT::DTOutput('table') #step2
)
server <- function( input, output, session ){
  output$plot <- renderPlot( { #step2 - adding placeholders
    ggplot()
  })
  output$table <- DT::renderDT({ #step2 - adding placeholders
    gapminder
  })
}
shinyApp( ui = ui, server = server )
```
```{r}
ui <- fluidPage(
  titlePanel( 'Life Expectation vs. GDP Per Capita' ),  #step3 - format the UI layout
  sidebarLayout( 
    sidebarPanel( 
      selectInput( 'continent', 'Select Continent', unique( gapminder$continent )), 
      sliderInput( 'year', 'Select Year', 1952, 2007, 1992, step = 5 ) 
      ),
    mainPanel( 
      tabsetPanel( 
        tabPanel( 'Plot', plotOutput( 'plot' ) ), 
        tabPanel( 'Table', DT::DTOutput('table') ) 
        ) 
      ) 
    ) 
)
server <- function( input, output, session ){   #step4 - update the outputs
  output$plot <- renderPlot( { #step2 - adding placeholders
    data <- gapminder %>%
      filter( year == input$year ) %>%
      filter( continent == input$continent )
    #print( data )
    ggplot( data, aes( x = gdpPercap, y = lifeExp )) +
      geom_point() +
      ylim( c( 50, 90 ) )
  })
  output$table <- DT::renderDT({ #step2 - adding placeholders
    gapminder %>%
      filter( year == input$year ) %>%
      filter( continent == input$continent )
  })
}
shinyApp( ui = ui, server = server )
```

```{r}
ui <- fluidPage(
  selectInput('greeting', 'Select greeting', choices = c("Hello", "Bonjour")),
  textInput( 'name', 'Enter your name', 'Kaelen' ),
  textOutput("q")
)

server <- function( input, output ){
  output$q <- renderText( {
    paste( input$greeting, ',', input$name )
  })
}
shinyApp( ui = ui, server = server )
```
```{r}
ui <- fluidPage(
  titlePanel( "Most Popular Names" ),
  sidebarLayout( sidebarPanel( selectInput('sex', 'Select sex', choices = c("M", "F")),
  sliderInput( "year", "Select year", 1880, 2017, 1 ) ),
  mainPanel( 
    tabsetPanel( 
      tabPanel( 'Plot', plotOutput( "plot" ) ), 
      tabPanel( 'Table', DT::DTOutput('table') ) ) ) )
)

server <- function(input, output, session) {
  output$plot <- renderPlot( {
      top_10_names <- babynames %>% 
      filter(sex == input$sex) %>% 
      filter(year == input$year ) %>% 
      top_n(10, prop)
      ggplot(top_10_names, aes(x = name, y = prop)) +
      geom_col()
  })
  output$table <- DT::renderDT({ #step2 - adding placeholders
    babynames %>% 
    filter(sex == input$sex) %>% 
    filter(year == input$year ) %>% 
    top_n(10, prop)
  })
}
shinyApp(ui = ui, server = server)
```



<br><br><br>